---
title: "CITS4009 Computational Data Analysis"
subtitle: "A sample Rmd file and explanation on IQR"
graphics: yes
author: Wei Liu (2023)
date: "Semester 2, 2023"

output:
  html_document:
    includes:
      before_body: style.html
    number_sections: true

---

# Introduction 

This sample Rmd file shows you how to do R code, data visualization,
alongside with textual explanation. To generate the html file, do the following
in R:

1. Bring in the `rmarkdown` library by typing: `library(rmarkdown)`

1. Change your working directory to where this `example_IQR.Rmd` file is.

1. Type: `render("example_IQR.Rmd", output_file="example_IQR.html")`

1. Open the generated `example_IQR.html` file in any web browser and view
   the output.

To generate the html file in RStudio is even simpler:

1. Open this `example_IQR.Rmd` file in RStudio.

2. Click the `knit` button. The generated `example_IQR.html` file should
   appear in your working directory and your default web browser
   should be launched to show the html file.

To quickly learn `rmarkdown`, see the following links:

* **R Markdown cheat sheet:** https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf

* **R Markdown from RStudio:** https://rmarkdown.rstudio.com/

# Defining a function

Firstly, we bring in the required library and define a function called
`plot.density.boxplot` that is needed several times in the code.
The function superimposes the boxplot and distribution of a given vector
of numbers on top of each other so that we can see the alignment of the
boxplot with the distribution.  This means that the boxplot will be in the
horizontal orientation, with the whiskers horizontal rather than vertical.

```{r size="tiny", eval=T, echo=T}
library(ggplot2)

plot.density.boxplot <- function(x) {
   # find out the highest peak of the distribution
   max.peak <- max(density(x)$y)
   # adjust the width of boxplot with respect to the max.peak value and
   # raise the boxplot a little bit so that the whiskers are slightly
   # above the x-axis.
   ggplot() + geom_density(aes(x=x), col="blue") +
      geom_boxplot(aes(x=x, y=0.1*max.peak), width=1.6*max.peak,
                   fill='yellow', alpha=0.5,
                   outlier.colour="red", outlier.size=6) +
      labs(title="Distribution and boxplot of the data", y="density")
}
```

# Experiments

## Experiment 1

Let's synthesize some data and call `boxplot.stats()` to get the
`Q0`, `Q1`, ...,  `Q4` values as mentioned in the lecture note.

```{r size="tiny", out.width="60%", collapse=T}
x <- rnorm(1000)
plot.density.boxplot(x)

q <- boxplot.stats(x)$stats # we call these numbers Q0,...,Q4
q
for (i in 0:4) 
   assign(paste('Q', i, sep=""), q[i+1])
```

The `for` loop above is the same as doing the following five assignment statements.
```{r eval=F, echo=T}
Q0 <- q[1]; Q1 <- q[2]; Q2 <- q[3]; Q3 <- q[4]; Q4 <- q[5]
```

In the code below, we compute the IQR and investigate how `Q0` and `Q4` are computed by the `boxplot.stats()` function.

```{r collapse=T}
# inter-quartile range
IQR <- Q3 - Q1

# Theoretically, the lower whisker Q0 should be at:
theoretical.Q0 <- Q1 - 1.5*IQR

cat("Theoretically, Q0 should be", theoretical.Q0, "\n")
cat("Q0 computed by boxplot.stats() is", Q0, "\n")
```

Instead of returning the theoretical `Q0` value using the
formula `Q1 - 1.5*IQR`, `boxplot.stats()` goes back to our
data (vector `x`) to find the smallest value in `x` that is
*just* larger than or equal to `theoretical.Q0`.

```{r collapse=T}
v0 <- min(x[x >= theoretical.Q0])
cat("The smallest value in x that is just larger than or equal to theoretical.Q0 is", v0, "\n")
```

We can see that variable `v0` is the same as the `Q0` value returned
by the `boxplot.stats()` function.  Similarly, instead of returning the
theoretical `Q4` value using the formula `Q3 + 1.5*IQR`, `boxplot.stats()`
goes back to our data (vector `x`) to find the largest value in `x`
that is *just* smaller than or equal to `theoretical.Q4`.

```{r collapse=T}
theoretical.Q4 <- Q3 + 1.5*IQR
v4 <- max(x[x <= theoretical.Q4])
cat("Theoretically, Q4 should be", theoretical.Q4, "\n")
cat("Q4 computed by boxplot.stats() is", Q4, "\n")
cat("The largest value in x that is just smaller than or equal to theoretical.Q4 is", v4, "\n")
```

The example above shows that `boxplot.stats()` tries to be *conservative*
so the lower and upper whiskers are always a little bit shorter than
the theoretical values. In this particular example, we have points
beyond the whiskers. That is, there are outliers in our data.
If our lower (or upper) whisker is *significantly* shorter
than the `1.5*IQR` value, does it mean that there are no outliers beyond
the quartile group 1 (or quartile group 4) region?

## Experiment 2

To answer the question above, let's consider the following bimodal
distribution which has a very wide IQR. Our data in this example
will therefore give a very large theoretical whisker length (`1.5*IQR`)
which will get cut back by our data.


```{r out.width="60%", collapse=T}
x <- sample(c(rnorm(100, mean=0.5), rnorm(100, mean=10, sd=0.5)))
plot.density.boxplot(x)

q <- boxplot.stats(x)$stats
q
for (i in 0:4) 
   assign(paste('Q', i, sep=""), q[i+1])

IQR <- Q3 - Q1
theoretical.Q0 <- Q1 - 1.5*IQR
v0 <- min(x[x >= theoretical.Q0])
cat("Theoretically, Q0 should be", theoretical.Q0, "\n")
cat("Q0 computed by boxplot.stats() is", Q0, "\n")
cat("The smallest value in x that is just larger than or equal to theoretical.Q0 is", v0, "\n")

theoretical.Q4 <- Q3 + 1.5*IQR
v4 <- max(x[x <= theoretical.Q4])
cat("Theoretically, Q4 should be", theoretical.Q4, "\n")
cat("Q4 computed by boxplot.stats() is", Q4, "\n")
cat("The largest value in x that is just smaller than or equal to theoretical.Q4 is", v4, "\n")
```

In this example, the two whiskers are much shorter than `1.5*IQR`.
Furthermore, a closer inspection reveals that `Q0` and `Q4` are also
the minimum and maximum values of our data.
```{r collapse=T}
cat("min(x) = ", min(x), "; max(x) = ", max(x))
```

Is it always the case that when the whiskers are much shorter than the
`1.5*IQR` value it means our data has no outliers?

## Experiment 3
To answer the question above, let's manually
contaminate the data with two outliers. In the two lines below, we
add two values `-50` and `50` to x. 

```{r out.width="60%", collapse=T}
x[length(x)+1] <- -50
x[length(x)+1] <- 50
plot.density.boxplot(x)
q <- boxplot.stats(x)$stats
q
```

Inspecting the `q` vectors from Experiment 2 and from this experiment,
we can see that the IQR regions for the data in the two experiments are
the same and the `Q0` and `Q4` values remain the same as well. Since the
value `-50` and `50` are outside the `Q1 - 1.5*IQR` and `Q3 + 1.5*IQR`
limits, the `boxplot.stat()` function picks the second number from the
sorted `x` vector for `Q0` and the second last number from the sorted
`x` vector for `Q4`:

```{r collapse=T}
sorted.x <- sort(x)
cat("The first four numbers in sorted x are:", sorted.x[1:4])
cat("Q0 returned by boxplot.stats() is:", q[1])

cat("The last four numbers in sorted x are:", rev(sorted.x)[4:1])
cat("Q4 returned by boxplot.stats() is:", q[5])
```

In the diagram above, we see that there is an outlier well beyond the tip
of each whisker.  

By *only* looking at the five numbers produced by `boxplot.stats()`,
we cannot conclude whether our data has outliers or not.  We can use the boxplot.out()

```{r collapse=T}
outliers <- boxplot.stats(x)$out

cat("The outliers in x are:", outliers)

```

* either inspect the boxplot diagram to *visually confirm* the absence/presence
  of outliers &mdash or use `boxplots.stats(x)$out`

* or inspect the minimum and maximum values of the data &mdash;
  if `Q0 = min(data)` then no outliers at the lower whisker end;
  if `Q4 = max(data)` then no outliers at the upper whisker end.
  

# Conclusion

This is a very detailed inspection of the five numbers returned
by `boxplot.stats()` and the importance of using IQR to identify
outliers/anomalies in our data. Outliers are not always wrong values that
should be removed. They can be just rare (but correct) measurements that
are well away from the central 50% (the IQR region) and the lowest or highest quartile of the population.

For the `rmarkdown` side, some useful options are listed below:

* `out.width` &mdash; for controlling the width of the output diagrams. This
  parameter won't affect the width of the textual outputs.
  The value should be a percentage value (relative to the width of
  the page) inside quotes. e.g., "60%".

* `collapse` &mdash; for collapsing the R code and R output together
  for compactness. This is a logical variable. The default
  value is `FALSE`.

* `eval` &mdash; for controlling whether the R code should be evaluated.
  Sometimes we want to display (`echo=T`) a snippet of R code without
  executing it (`eval=F`).  The default value is `TRUE`.

* `echo` &mdash; for controlling whether the R code should appear in the output
  html file. Sometimes we want the R code to be run silently with
  the output displayed but not showing the code (`echo=F`). The default
  value is `TRUE`.

* `warning` &mdash; for allowing/suppressing warning messages from R to appear
  in the output html file. This is a logical variable.
  `warning=FALSE` means to suppress warning messages.
  
<br><br>
